Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2013-03-08 HRUIonGPU/jrnl_article.tex

\begin{abstract}
%\boldmath
Capon beamforming is associated with a high computational complexity, which limits its use as a real-time array processor in many applications. In this paper we present an implement of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Most of this work focus on the improved lateral resolution and contrast obtained with the Capon beamformer. However, some interesting trade-offs has also been introduced. In \cite{Synnevag2009}, Synnev\aa{}g et al. explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration. Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}. In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.

Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image (Fig 9. in \cite{Synnevag2009}). Although complex arithmetic requires more operations per sample, complex (I/Q or baseband) data can be sampled at a lower rate, thus minimizing the number of samples needed per image. All results in this paper are reported with the use of baseband data, hence where the sampling frequency is close to the system bandwidth.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a real-time requirement less strict than medical ultrasound imaging, which means that most array configurations can be handled in real-time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is a cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen et al. \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and how it has been implemented on the GPU, is presented in Section \ref{sec:bs}. Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.  Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
\caption{Capon beamforming. Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] =  f(\vec{x}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
\end{align}
A weight proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. Since medical ultrasound imaging is an active, short-pulsed, system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies. As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2011}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores and logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.

The CUDA compute model is based on execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (32x32). Threads inside a block supports low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SMs), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of near-core memory and registers. Hence, to achieve maximum GPU occupancy (having the maximum number of threads residing on each SM), each thread has to be as light on register and memory consumption as possible. It is also important that each thread performs enough instructions per transferred byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency. Reads and writes to and from GPU DRAM, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.

\begin{figure}
\centerline{\includegraphics[width=3.4in]{gfx/gpu_layout_vertical_lr.png}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

\section{Parallel Capon}\label{sec:meth}
If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with (\ref{eq:b}) and (\ref{eq:diag}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}. However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$ (if $L=M/2$). Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

Diagonal loading as a constant value can be added with negligible cost. For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to GPU, is avoided.

For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch. The solver is available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increasing the weight vector to length $M$. Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to  GPU DRAM. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

>>>>

\begin{abstract}
%\boldmath
Capon beamforming is associated with a high computational complexity, which limits its use as a real-time method in many applications. In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging are shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer. However, some interesting trade-offs have also been introduced. In \cite{Synnevag2009}, Synnev\aa{}g et al. explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration. Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}. In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.

Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. However, the Capon beamformer is meant to work on complex data. Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image (Fig 9. in \cite{Synnevag2009}). Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced. The number of samples needed per image is therefore minimized. All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is close to the system bandwidth.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen et al. \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}. Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.  Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
\caption{Capon beamforming. Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] =  f(\vec{x}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
\end{align}
A weight proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies. As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.

The CUDA compute model is based on the execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block ($32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of near-core, or local, memory and registers. Hence, to achieve maximum GPU occupancy (having the maximum number of threads residing on each SM), each thread has to be as light on register and memory consumption as possible. It is also important that each thread performs enough instructions per transferred byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency. Reads and writes to and from global GPU memory, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

\section{Parallel Capon}\label{sec:meth}
If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with (\ref{eq:b}) and (\ref{eq:diag}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}. However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$. Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.

@@

For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.

For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch. The solver is available online for registered CUDA developers \cite{Nvidiaa}.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increasing the weight vector to length $M$. Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
